Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-04-28 overleaf/compiler.tex
Respond to motion translation comments

% SIMD representation
Vectorization is handled with utility functions to manage accessing and updating slices of arrays.  All SIMD values are stored in non-vectorized form as 1-dimensional \texttt{std::vector}s in row-major order. \ana{Does MOTION support just 1-dim SIMD arrays? If yes, mention it here.} Whenever a SIMD value is used in an expression, the utility function \texttt{vectorized\_access()} takes the multi-dimentional representation of a SIMD value, along with the size of each dimension and the requested slice's indices, and converts that slice to a MOTION SIMD value. \ana{Add the reference to the ADD example figure here.} Because MOTION supports SIMD operations using the same C++ operators as non-SIMD operations, we do not need to perform any other transformations to the expression. \ana{Is it correct to say that the operators are overloaded? If yes, use that term?} Therefore, the translation of an expression containing SIMD values is identical to that of expressions without SIMD values.

Similarly, the \texttt{vectorized\_assign()} function assigns a (potentially SIMD) value to a slice of a vectorized array.  This operation cannot be done with a simple subscript as SIMD assignments will update a range of values in the non-vectorized representation.
\ben{Is the next sentence needed?  They can always check the implementation of vectorized\_assign() if they want to see specifics.}
The function takes the array being assigned to as a reference to update it in place so that any existing values are not lost.  SIMD values are never directly assigned to non-SIMD values. \ana{Yes, this is implementation-specific, you can remove.}

@@

\ben{The compiler and C++ internally uses ``lift'' as the terminology for raise\_dim(), but the paper does not :(...} \ana{Yes, that's true. To make matters worse, most of the explanation on this comes in the next section. But I'll leave as is for now.}
\paragraph{Reshaping and raising dimensions:}
Raising the dimensions of a scalar or array uses the \texttt{lift()} utility function which takes a lambda for the raised expression and the dimensions of the output.  The lambda accepts an array of integers representing the index along each of the dimensions being raised, and the translation of the expression which is being raised replaces each of the dimension index variables with the relevant subscript of this array.  There is also a special case of the \texttt{lift()} function which occurs when we are raising an array.  This case only occurs when the dimensionality of the array is equal to the first dimensions of the array (any dimension which is being lifted to which is not in the array is after every dimension in the array) \ben{I don't know how to phrase the previous sentence more concisely}.  In this case, the array is simply extended along the new dimensions (if the array is already correctly sized it is returned as is). \ana{Add the ref. to the figure. Also, try to rewrite. I understand what the paragraph is saying but it is too technical and does not convey the higher-level goal of these. Try first to state what the goal is, and then talk about the details such as the lambda funcs.}

@@

% drop_dim() implementation
\paragraph{Dropping dimensions:}
Dropping dimensions use the \texttt{drop\_dim()} and \texttt{drop\_dim\_monoreturn()} utility functions.  They function identically but the latter returns a scalar for the case when the final dimension of an array is dropped.  These functions take the non-vectorized representation of an array, along with the dimensions of that array, and return the array with the final dimension dropped.
\ben{drop\_dim() taking the true branch of phi node values doesn't need to be covered here, right?} \ana{Right, I wouldn't add it here.}

% Shared/plaintext conversions
\ben{Ishaq or Vassilis: please check over this paragraph... I don't want to make some mistake about MOTION's capabilities.}
Currently, our compiler only supports the \texttt{Bmr} and \texttt{BooleanGMW} protocols as MOTION does not implement all operations for other protocols \ben{Should that go here?} \ana{Yes, I think it's good}.  MOTION does not support publicly-known constants for these protocols, so all conversions from plaintext values to shares are performed by providing the plaintext value as a shared input from party 0.  Due to this limitation, our translation to MOTION code attempts to minimize the number of conversions into a plaintext value. \ana{Typo: conversions from plaintext values, right?} This is accomplished by creating a shared copy of each plaintext variable and updating that copy in lock-step with the plaintext variable.  Since variables are often initialized to a common constant value (e.g. 0), this approach decreases the number of input gates by only creating a shared input for each initialization constant.  Loop counters must still be converted to a shared value on each iteration that they are used, however we only generate this conversion when necessary, i.e., when the counter flows to a shared computation. This is to prevent unnecessary increase in the number of input gates.

>>>>

% SIMD representation
Vectorization is handled with utility functions to manage accessing and updating slices of arrays.  All SIMD values are stored in non-vectorized form as 1-dimensional \texttt{std::vector}s in row-major order. \ana{Does MOTION support just 1-dim SIMD arrays? If yes, mention it here.} \ben{MOTION does only support 1-dim SIMD arrays, but since we use helper functions to create and access SIMD arrays this restriction doesn't apply to our representation. There is no reason why these arrays need to be 1-dimensional other than that it slightly simplifies the code generation and avoids some nasty templates on the utility functions.} Whenever a SIMD value is used in an expression, the utility function \texttt{vectorized\_access()} takes the multi-dimentional representation of a SIMD value, along with the size of each dimension and the requested slice's indices, and converts that slice to a MOTION SIMD value. \ana{Add the reference to the ADD example figure here.} Because MOTION supports SIMD operations using the same C++ operators as non-SIMD operations, we do not need to perform any other transformations to the expression. \ana{Is it correct to say that the operators are overloaded? If yes, use that term?} \ben{I don't think that's the right term since SIMD and scalar MOTION values use the same C++ class, so the operator is the same.} Therefore, the translation of an expression containing SIMD values is identical to that of expressions without SIMD values.

Similarly, the \texttt{vectorized\_assign()} function assigns a (potentially SIMD) value to a slice of a vectorized array.  This operation cannot be done with a simple subscript as SIMD assignments will update a range of values in the non-vectorized representation.
%\ben{Is the next sentence needed?  They can always check the implementation of vectorized\_assign() if they want to see specifics.}
%The function takes the array being assigned to as a reference to update it in place so that any existing values are not lost.  SIMD values are never directly assigned to non-SIMD values. \ana{Yes, this is implementation-specific, you can remove.}

@@

\ben{The compiler and C++ internally uses ``lift'' as the terminology for raise\_dim(), but the paper does not :(...} \ana{Yes, that's true. To make matters worse, most of the explanation on this comes in the next section. But I'll leave as is for now.}
\paragraph{Reshaping and raising dimensions:}
Raising the dimensions of a scalar or array uses the \texttt{lift()} utility function which takes a lambda for the raised expression and the dimensions of the output.  This function evaluates the expression for each permutation of indices along the dimensions and returns the resulting array in row-major order.  The lambda accepts an array of integers representing the index along each of the dimensions being raised, and the translation of the expression which is being raised replaces each of the dimension index variables with the relevant subscript of this array.  There is also a special case of the \texttt{lift()} function which occurs when we are raising an array.  In this case, instead of concatenating the array for each index, we extend the array along all dimensions being raised which are not present in the array already.  For example, when raising an array with dimensions $N \times M$ to an array with dimensions $N \times M \times D$, the input array will simply be extended along the $D$ dimension.  If input array is already correctly sized it will be returned as-is.
%This case only occurs when the dimensionality of the array is equal to the first dimensions of the array (any dimension which is being lifted to which is not in the array is after every dimension in the array) \ben{I don't know how to phrase the previous sentence more concisely}.  In this case, the array is simply extended along the new dimensions (if the array is already correctly sized it is returned as is).
\ana{Add the ref. to the figure. Also, try to rewrite. I understand what the paragraph is saying but it is too technical and does not convey the higher-level goal of these. Try first to state what the goal is, and then talk about the details such as the lambda funcs.} \ben{Rewrote the above paragraph to be less confusing/implementation-specific, please review.}

@@

% drop_dim() implementation
\paragraph{Dropping dimensions:}
Dropping dimensions use the \texttt{drop\_dim()} and \texttt{drop\_dim\_monoreturn()} utility functions.  They function identically but the latter returns a scalar for the case when the final dimension of an array is dropped.  These functions take the non-vectorized representation of an array, along with the dimensions of that array, and return the array with the final dimension dropped.

% Shared/plaintext conversions
\ben{Ishaq or Vassilis: please check over this paragraph... I don't want to make some mistake about MOTION's capabilities.}
Currently, our compiler only supports the \texttt{Bmr} and \texttt{BooleanGMW} protocols as MOTION does not implement all operations for other protocols.  MOTION does not support publicly-known constants for these protocols, so all conversions from plaintext values to shares are performed by providing the plaintext value as a shared input from party 0.  Due to this limitation, our translation to MOTION code attempts to minimize the number of conversions from a plaintext value. This is accomplished by creating a shared copy of each plaintext variable and updating that copy in lock-step with the plaintext variable.  Since variables are often initialized to a common constant value (e.g. 0), this approach decreases the number of input gates by only creating a shared input for each initialization constant.  Loop counters must still be converted to a shared value on each iteration that they are used, however we only generate this conversion when necessary, i.e., when the counter flows to a shared computation. This is to prevent unnecessary increase in the number of input gates.
